Reasonably Programmable Literal Notation

### 2018-06-25 icfp18/syntax-icfp18.tex
fixed a couple of line number omissions and dropped unnecessary distinction between re_expr and proto_expr

\begin{figure}
\begin{lstlisting}[deletekeywords={spliced}]
  module Segment : { 
    type t = {start_pos: int, end_pos: int};
    let read_to : string => Lexing.lexbuf => Segment.t; };
  exception ExpansionError({msg: string, loc: option(Segment.t)});
  module ProtoExpr : { 
    type t = Parsetree.expression; /* from the OCaml compiler library */
    let spliced : Segment.t => Parsetree.core_type => t; };
  notation $re_expr at Parsetree.expression { /* ... (see text) ... */ };
  notation $proto_expr at ProtoExpr.t = $re_expr;
\end{lstlisting}
\vspace{-2px}
\caption{A fragment of the \li{Relit} module's signature relevant to expression TLM providers.}
\label{fig:relit-util}
\vspace{-5px}
\end{figure}
    % let match_spliced : t => option(Segment.t, ProtoTyp.t);

@@

The datatypes in OCaml's \li{Parsetree} module are necessarily intricate, given the sophistication of the OCaml system, so constructing expression encodings manually is generally too unwieldy. Addressing this class of problem is, of course, the motivation for this very paper, so it is only natural that we define TLMs  for working with OCaml parse trees using familiar surface syntax, extended with splice forms. In the literature, literal notation for the host language's own parse trees is called \emph{quasiquotation}, and splicing is referred to as \emph{unquotation} or \emph{antiquotation} \cite{Bawd99a,shabalin2013quasiquotes,mainland2007s}. The TLM \li{$re_expr}  specified on Line 10 of Fig.~\ref{fig:relit-util} provides quasiquotation for OCaml parse trees using Reason's surface syntax. The accompanying technical report describes its implementation. For clarity, the \li{$proto_expr} TLM is defined as a synonym for \li{$re_expr} on Line~12 of Fig.~\ref{fig:relit-util} (the two TLMs may diverge in the future). We apply this TLM many times in the semantic actions in Fig.~\ref{fig:regex-parser}. Notice, however, that the generalized literal forms in Fig.~\ref{fig:regex-parser} are not each prefixed by \li{$proto_expr}. Instead, we use the \li{open notation} directive on Line 2, which implicitly applies \li{$proto_expr} to all unadorned generalized literal forms in its scope.

To support antiquotation, the \li{$re_expr} TLM repurposes TLM application and generalized literal forms, as can be observed throughout Fig.~\ref{fig:regex-parser}. In particular, parse tree splicing is supported by unadorned generalized literal forms, e.g. on Line~23 where we implement the regex notation \li{r+} in terms of sequencing and star, taking care to bind \li{r} to an expansion-internal variable---also \li{r} but distinguished by the segmentation---to avoid double evaluation. String splicing (which converts the spliced string expression into a parse tree of string constant form) is performed by repurposing the TLM name \li{$s} (Line~16). There is also notation for references to spliced expressions that repurposes the TLM name \li{$spliced}, as seen on Lines 28-32, where the regex splicing logic is implemented.

TLM-related forms are available to be repurposed in this way by \li{$proto_expr} because there is no reasonable way for \li{$re_expr} to support antiquotation across notational boundaries. For example, if \li{$re_expr} had instead used some other antiquotation notation, e.g. \li{^(}$e$\li{)}, then consider this example:
\begin{lstlisting}[numbers=none]
  $proto_expr `( SURLleEURLSURLt y = ^(EURLr1SURL); $m `(...{^(r2)}...)`EURL )`
\end{lstlisting}
The sequence \li{^(r1)} is in expression position, so antiquotation occurs. However, there is no way for \li{$re_expr} to know how the TLM that \li{$m} will eventually refer to will parse the sequence \li{^(r2)}---it may end up in a spliced expression, or it may have some other meaning---so it cannot perform antiquotation. In other words, TLM expansion is fundamentally ``outside-in''.  Notice that the segmentation does communicate the fact that \li{r2} has not been antiquoted (it remains lavender), but to avoid confusion, we decided not to support the generation of TLM applications via \li{$re_expr}.

@@

Although \li{ocamllex} and Menhir are powerful tools, they are not right for every parsing job. For example, we might want to use a different parser generator, a parser combinator library \cite{Hutton1992d}, or post-process the result of calling an existing parser to produce a corresponding proto-expansion. Fortunately, it is easy to bypass \li{ocamllex} and Menhir. 
%The most obvious method is to define a trivial lexer that reads the entire literal body into a single token, and a trivial parser with one semantic action. In fact, this is not actually necessary because the system does not interface with \li{ocamllex} and Menhir directly. Rather, it is the job of the user's build system to make sure that by the time the compiler is invoked, these tools have already been invoked as necessary to produce ordinary OCaml modules satisfying the expected interface. It is just as well to implement this interface manually. 
% the \li{Relit} library includes a trivial lexer, \li{Relit.TrivLexer}, and a few functors that take care of the necessary boilerplate in situations where we simply want to define the parser as a function of type \li{string => Relit.ProtoExpr.t}. 
%For example, the accompanying technical report briefly describes how \li{$re_expr}, described above, can be expressed in terms of the Reason parser, a parse tree serializer and a simple tree transformation. 
Let us consider one substantial example of this approach here: the \li{$html} TLM, shown being applied in Fig.~\ref{fig:nesting-example}, is implemented using an existing, ``production grade'' HTML parsing library, \li{Markup.ml} \cite{markupml}, together with a simple post-processing step as outlined in Fig.~\ref{fig:html-case-study}.

@@

The \li{ProtoHtml.parse} function, outlined on Lines 8-9, generates a value of the type \li{ProtoHtml.t} defined on Lines 4-7 and so named because it is similar to \li{Html.t} but distinguishes spliced segments, like a proto-expression. We omit the definition of \li{ProtoHtml.parse} but conceptually, it starts by creating a character stream from the literal body, then from that creates a \li{Markup.ml} HTML content stream, then transforms that to a stream that emits the signals from the HTML content stream until it sees a splice start tag (which is reported as a syntax error relative to the HTML standard), at which point it calls \li{Segment.read_to}, described earlier, to advance the original character stream until it recognizes the closing tag, emitting the generated segment and then returning control to the HTML content stream. Finally, it folds over this stream to produce the result of type \li{ProtoHtml.t}.

@@

\begin{figure}
\begin{subfigure}[t]{0.4\textwidth}
\begin{lstlisting}
  module Map = { 
    module type S = {
      type key;
      type t('a);
      let empty : t('a);
      let add : 
        key => 'a => t('a) => t('a);
      /* ... */
    };
  };
\end{lstlisting}
\vspace{-7px}
\caption{The \li{Map.S} signature}
\label{fig:map-sig}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.55\textwidth}
\begin{subfigure}[t]{\textwidth}
\begin{lstlisting}
  notation $map(M : Map.S, type a) at M.t(a) {
    lexer MapLexer parser MapParser 
    in package map_parser;
  }
\end{lstlisting}
\vspace{-8px}
\caption{The parametric TLM \li{$map}}
\vspace{5px}
\label{fig:map-tlm}
\end{subfigure}
\begin{subfigure}[b]{\textwidth}
\begin{lstlisting}
  notation $stringmap = $map(Misc.StringMap);
  let make_request id req v = $stringmap
    `( "session_id" SURL->EURL string_of_int(id)SURL,EURL
                req SURL->EURL v )`;
\end{lstlisting}
\vspace{-4px}
\caption{Applying \li{$map}}
\label{fig:map-ap}
\end{subfigure}
\end{subfigure}
\vspace{4px}
\caption{Literal notation for all finite map implementations using parametric TLMs.}
\vspace{-5px}
\label{fig:map-fig}
\end{figure}

@@

An alternative point-of-view in ML is to treat the \li{dependencies} clause as fundamental and achieve parameterization using module functions (functors), e.g. we could express \li{$map} as follows:
\begin{lstlisting}[numbers=none]
  module MapNotation = (M : Map.S, A : { type a; }) => {
    notation $map at M.t(A.a) { 
      lexer MapLexer parser MapParser in package map_parser;
      dependencies { module M = M; type a = A.a; } } }
\end{lstlisting}
% The conveniences related to type parameter inference are not available when parameterization is encoded in this way.

@@

% \begin{subequations}[resume]
% \begin{equation*}\label{rule:expandsU-syntax}
% \inferrule{
%   \istypeU{\Delta}{\tau}\\
%   \expandsU{\emptyset}{\emptyset}{\emptyset}{\ueparse}{\eparse}{\aparr{\tBody}{\tParseResultExp}}\\\\
%   \expandsU{\Delta}{\Gamma}{\uPsi, \xuetsmbnd{\tsmv}{\tau}{\eparse}}{\ue}{e}{\tau'}
% }{
%   \expandsUX{\audefuetsm{\tau}{\ueparse}{\tsmv}{\ue}}{e}{\tau'}
% }
% \end{equation*}
\begin{figure*}
{\small\begin{mathpar}
\inferrule[ee-def-setlm]{
  \expandsTU{\uDelta}{\utau}{\tau}\\\\
  \hastypeU{\emptyset}{\emptyset}{\eparse}{\aparr{\tBody}{\tParseResultExp}}\\
  \expandsUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{\uedep}{\edep}{\taudep}\\\\
  \uGamma = \uGG{\uG}{\Gamma}\\
  \expandsUP{\uDelta}{\uGG{\uG}{\Gamma, x : \taudep}}{\uPsi, \uShyp{\tsmv}{x}{\tau}{\eparse}}{\uPhi}{\ue}{e}{\tau'}\\\\
  e_\text{defn} = \aeap{\aelam{\taudep}{x}{e}}{\edep}
}{
  \expandsUPX{\uesyntaxqr{\tsmv}{\utau}{\eparse}{\ue}}{e_\text{defn}}{\tau'}
}
\end{mathpar}}
{\small\begin{mathpar}
\inferrule[ee-def-sptlm]{
  \expandsTU{\uDelta}{\utau}{\tau}\\
  \hastypeU{\emptyset}{\emptyset}{\eparse}{\aparr{\tBody}{\tParseResultPat}}\\\\
  \expandsUP{\uDelta}{\uGamma}{\uPsi}{\uPhi, \uPhyp{\tsmv}{\_}{\tau}{\eparse}}{\ue}{e}{\tau'}
}{
  \expandsUPX{\usyntaxup{\tsmv}{\utau}{\eparse}{\ue}}{e}{\tau'}
}
\end{mathpar}}
\vspace{-4px}
\caption{The typed expansion rules for expression and pattern TLM definitions.}
\label{fig:ee-def-setlm}
\vspace{-5px}
\end{figure*}
% \end{subequations}

>>>>

\begin{figure}
\begin{lstlisting}[deletekeywords={spliced}]
  module Segment : { 
    type t = {start_pos: int, end_pos: int};
    let read_to : string => Lexing.lexbuf => Segment.t; };
  exception ExpansionError({msg: string, loc: option(Segment.t)});
  module ProtoExpr : { 
    type t = Parsetree.expression; /* from the OCaml compiler library */
    let spliced : Segment.t => Parsetree.core_type => t; };
  notation $proto_expr at ProtoExpr.t { /* ... (see text) ... */ };
\end{lstlisting}
\vspace{-2px}
\caption{A fragment of the \li{Relit} module's signature relevant to expression TLM providers.}
\label{fig:relit-util}
\vspace{-5px}
\end{figure}
    % let match_spliced : t => option(Segment.t, ProtoTyp.t);

@@

The datatypes in OCaml's \li{Parsetree} module are necessarily intricate, given the sophistication of the OCaml system, so constructing expression encodings manually is generally too unwieldy. Addressing this class of problem is, of course, the motivation for this very paper, so it is only natural that we define TLMs  for working with OCaml parse trees using familiar surface syntax, extended with splice forms. In the literature, literal notation for the host language's own parse trees is called \emph{quasiquotation}, and splicing is referred to as \emph{unquotation} or \emph{antiquotation} \cite{Bawd99a,shabalin2013quasiquotes,mainland2007s}. The TLM \li{$proto_expr}  specified on Line 8 of Fig.~\ref{fig:relit-util} provides quasiquotation for OCaml parse trees using Reason's surface syntax. The accompanying technical report describes its implementation. We apply this TLM many times in the semantic actions in Fig.~\ref{fig:regex-parser}. Notice, however, that the generalized literal forms in Fig.~\ref{fig:regex-parser} are not each prefixed by \li{$proto_expr}. Instead, we use the \li{open notation} directive on Line 2, which implicitly applies \li{$proto_expr} to all unadorned generalized literal forms in its scope.

To support antiquotation, the \li{$proto_expr} TLM repurposes TLM application and generalized literal forms, as can be observed throughout Fig.~\ref{fig:regex-parser}. In particular, parse tree splicing is supported by unadorned generalized literal forms, e.g. on Line~23 where we implement the regex notation \li{r+} in terms of sequencing and star, taking care to bind \li{r} to an expansion-internal variable---also \li{r} but distinguished by the segmentation---to avoid double evaluation. String splicing (which converts the spliced string expression into a parse tree of string constant form) is performed by repurposing the TLM name \li{$s} (Line~16). There is also notation for references to spliced expressions that repurposes the TLM name \li{$spliced}, as seen on Lines 28-32, where the regex splicing logic is implemented.

TLM-related forms are available to be repurposed in this way by \li{$proto_expr} because there is no reasonable way for \li{$proto_expr} to support antiquotation across notational boundaries. For example, if \li{$proto_expr} had instead used \li{^(}$e$\li{)} for antiquotation, then consider this example:
\begin{lstlisting}[numbers=none]
  $proto_expr `( SURLleEURLSURLt y = ^(EURLr1SURL); $m `(...{^(r2)}...)`EURL )`
\end{lstlisting}
The sequence \li{^(r1)} is in expression position, so antiquotation occurs. However, there is no way for \li{$proto_expr} to know how the TLM that \li{$m} will eventually refer to will parse the sequence \li{^(r2)}---it may end up in a spliced expression, or it may have some other meaning---so it cannot perform antiquotation. In other words, TLM expansion is fundamentally ``outside-in''.  Notice that the segmentation does communicate the fact that \li{r2} has not been antiquoted (it remains lavender), but to avoid confusion, we decided not to support the generation of TLM applications via \li{$proto_expr}.

@@

Although \li{ocamllex} and Menhir are powerful tools, they are not right for every parsing job. For example, we might want to use a different parser generator, a parser combinator library \cite{Hutton1992d}, or post-process the result of calling an existing parser to produce a corresponding proto-expansion. Fortunately, it is easy to bypass \li{ocamllex} and Menhir. 
%The most obvious method is to define a trivial lexer that reads the entire literal body into a single token, and a trivial parser with one semantic action. In fact, this is not actually necessary because the system does not interface with \li{ocamllex} and Menhir directly. Rather, it is the job of the user's build system to make sure that by the time the compiler is invoked, these tools have already been invoked as necessary to produce ordinary OCaml modules satisfying the expected interface. It is just as well to implement this interface manually. 
% the \li{Relit} library includes a trivial lexer, \li{Relit.TrivLexer}, and a few functors that take care of the necessary boilerplate in situations where we simply want to define the parser as a function of type \li{string => Relit.ProtoExpr.t}. 
%For example, the accompanying technical report briefly describes how \li{$proto_expr}, described above, can be expressed in terms of the Reason parser, a parse tree serializer and a simple tree transformation. 
Let us consider one substantial example of this approach here: the \li{$html} TLM, shown being applied in Fig.~\ref{fig:nesting-example}, is implemented using an existing, ``production grade'' HTML parsing library, \li{Markup.ml} \cite{markupml}, together with a simple post-processing step as outlined in Fig.~\ref{fig:html-case-study}.

@@

The \li{ProtoHtml.parse} function, elided on Lines 8-9, generates a value of the type \li{ProtoHtml.t} defined on Lines 4-7 and so named because it is similar to \li{Html.t} but distinguishes spliced segments, like a proto-expression. We omit the definition of \li{ProtoHtml.parse} but conceptually, it starts by creating a character stream from the literal body, then from that creates a \li{Markup.ml} HTML content stream, then transforms that to a stream that emits the signals from the HTML content stream until it sees a splice start tag (which is reported as a syntax error relative to the HTML standard), at which point it calls \li{Segment.read_to}, described earlier, to advance the original character stream until it recognizes the closing tag, emitting the generated segment and then returning control to the HTML content stream. Finally, it folds over this stream to produce the result of type \li{ProtoHtml.t}.

@@

\begin{figure}
\begin{subfigure}[t]{0.4\textwidth}
\begin{lstlisting}
  module Map = { 
    module type S = {
      type key;
      type t('a);
      let empty : t('a);
      let add : 
        key => 'a => t('a) => t('a);
      /* ... */
    };
  };
\end{lstlisting}
\vspace{-7px}
\caption{The \li{Map.S} signature}
\label{fig:map-sig}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.55\textwidth}
\begin{subfigure}[t]{\textwidth}
\begin{lstlisting}
  notation $map(M : Map.S, type a) at M.t(a) {
    lexer MapLexer parser MapParser 
    in package map_parser;
  };
\end{lstlisting}
\vspace{-8px}
\caption{The parametric TLM \li{$map}}
\vspace{5px}
\label{fig:map-tlm}
\end{subfigure}
\begin{subfigure}[b]{\textwidth}
\begin{lstlisting}
  notation $stringmap = $map(Misc.StringMap);
  let make_request id req v = $stringmap
    `( "session_id" SURL->EURL string_of_int(id)SURL,EURL
                req SURL->EURL v )`;
\end{lstlisting}
\vspace{-4px}
\caption{Applying \li{$map}}
\label{fig:map-ap}
\end{subfigure}
\end{subfigure}
\vspace{4px}
\caption{Literal notation for all finite map implementations using parametric TLMs.}
\vspace{-5px}
\label{fig:map-fig}
\end{figure}

@@

An alternative point-of-view in ML is to treat the \li{dependencies} clause as fundamental and achieve parameterization using module functions (functors), e.g. we could express \li{$map} as follows:
\begin{lstlisting}[numbers=none]
  module MapNotation = (M : Map.S, A : { type a; }) => {
    notation $map at M.t(A.a) { 
      lexer MapLexer parser MapParser in package map_parser;
      dependencies { module M = M; type a = A.a; } 
    };
  };
\end{lstlisting}
% The conveniences related to type parameter inference are not available when parameterization is encoded in this way.

@@

% \begin{subequations}[resume]
% \begin{equation*}\label{rule:expandsU-syntax}
% \inferrule{
%   \istypeU{\Delta}{\tau}\\
%   \expandsU{\emptyset}{\emptyset}{\emptyset}{\ueparse}{\eparse}{\aparr{\tBody}{\tParseResultExp}}\\\\
%   \expandsU{\Delta}{\Gamma}{\uPsi, \xuetsmbnd{\tsmv}{\tau}{\eparse}}{\ue}{e}{\tau'}
% }{
%   \expandsUX{\audefuetsm{\tau}{\ueparse}{\tsmv}{\ue}}{e}{\tau'}
% }
% \end{equation*}
\begin{figure}
{\small\begin{mathpar}
\inferrule[ee-def-setlm]{
  \expandsTU{\uDelta}{\utau}{\tau}\\\\
  \hastypeU{\emptyset}{\emptyset}{\eparse}{\aparr{\tBody}{\tParseResultExp}}\\
  \expandsUP{\uDelta}{\uGamma}{\uPsi}{\uPhi}{\uedep}{\edep}{\taudep}\\\\
  \uGamma = \uGG{\uG}{\Gamma}\\
  \expandsUP{\uDelta}{\uGG{\uG}{\Gamma, x : \taudep}}{\uPsi, \uShyp{\tsmv}{x}{\tau}{\eparse}}{\uPhi}{\ue}{e}{\tau'}\\\\
  e_\text{defn} = \aeap{\aelam{\taudep}{x}{e}}{\edep}
}{
  \expandsUPX{\uesyntaxqr{\tsmv}{\utau}{\eparse}{\ue}}{e_\text{defn}}{\tau'}
}
\end{mathpar}}
{\small\begin{mathpar}
\inferrule[ee-def-sptlm]{
  \expandsTU{\uDelta}{\utau}{\tau}\\
  \hastypeU{\emptyset}{\emptyset}{\eparse}{\aparr{\tBody}{\tParseResultPat}}\\\\
  \expandsUP{\uDelta}{\uGamma}{\uPsi}{\uPhi, \uPhyp{\tsmv}{\_}{\tau}{\eparse}}{\ue}{e}{\tau'}
}{
  \expandsUPX{\usyntaxup{\tsmv}{\utau}{\eparse}{\ue}}{e}{\tau'}
}
\end{mathpar}}
\caption{The typed expansion rules for expression and pattern TLM definitions.}
\label{fig:ee-def-setlm}
\vspace{-5px}
\end{figure}
% \end{subequations}
